Notes on \\ Projected Knowledge Compilation \\ \today

### 2023-08-01 doc/pkc-notes.tex
Minor corrections

  Let $\varset$ denote a set of Boolean variables, and let $\assign$
  be an \emph{assignment} of truth values to some subset of the
  variables, where $0$ denotes false and $1$ denotes true, i.e.,
  $\assign \colon Y \rightarrow \{0,1\}$ for some $Y  \subseteq \varset$.
We say the assignment is \emph{total with respect to a set of variables} $Y$, when it assigns a value to every $x \in Y$, and that it is simply \emph{total} when it is total with respect to $\varset$.
  The set of all possible total assignments with respect to a set of variables $Y$ is denoted $\uassign(Y)$.

@@

\begin{definition}[Boolean Formulas]
  The set of Boolean formulas is defined recursively.  Each
  formula $\theta$ has an associated \emph{dependency set}
  $\dependencyset(\theta)  \subseteq \varset$, and a set of \emph{models} $\modelset(\theta)$,
  consisting of total assignments that satisfy the formula:
  \begin{enumerate}
  \item Constant formulas $\nil$ (logical falsehood) and $\tautology$ (tautology) are Boolean formulas,
    with $\dependencyset(\nil) = \dependencyset(\tautology) = \emptyset$, with $\modelset(\nil) = \emptyset$, and with $\modelset(\tautology) = \uassign(\varset)$.
  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
    and $\modelset(x) = \{\assign \in \uassign(\varset) \mid  \assign(x)=1\}$.
  \item For formula $\theta$, its \emph{negation}, written $\boolnot \theta$ is a Boolean formula,
    with $\dependencyset(\boolnot \theta) = \dependencyset(\theta)$ and $\modelset(\boolnot \theta) = \uassign(\varset) - \modelset(\theta)$.
  \item For formulas $\theta_1, \theta_2, \ldots, \theta_k$, their \emph{product} $\theta = \bigwedge_{1 \leq i \leq k} \theta_i$ is a Boolean formula, with
      $\dependencyset(\theta) = \bigcup_{1 \leq i \leq k} \dependencyset(\theta_i)$ and
      $\modelset(\theta) = \bigcap_{1 \leq i \leq k} \modelset(\theta_i)$.
  \item For formulas $\theta_1, \theta_2, \ldots, \theta_k$, their \emph{sum} $\theta = \bigvee_{1 \leq i \leq k} \theta_i$ is a Boolean formula, with
      $\dependencyset(\theta) = \bigcup_{1 \leq i \leq k} \dependencyset(\theta_i)$ and
      $\modelset(\theta) = \bigcup_{1 \leq i \leq k} \modelset(\theta_i)$.
  \end{enumerate}
\label{def:boolean}
\end{definition}

@@

  Many tools for reasoning about Boolean formulas operate on formulas in \emph{conjunctive normal form}, defined as follows.
  Formula $\theta$ is in  \emph{negation normal form} when negation is applied only to variables.  A
  formula is in \emph{conjunctive normal form} (CNF) when i) it is in
  negation normal form, and ii) sum is applied only to literals.

  A CNF
  formula can be represented as a set of \emph{clauses}, each of which is a
  set of literals.  Each clause represents the sum of the
  literals, and the formula is the product of its clauses.  We use
  set notation to reference the clauses within a formula and the
  literals within a clause.  A clause consisting of a single literal is referred to as a \emph{unit} clause and the literal as a \emph{unit} literal.
This literal must be assigned value $1$ by any satisfying assignment of the formula.  We the symbol $\phi$ to denote a CNF formula.

@@

We define the following operations for transforming clausal formula $\phi$ into a new clausal formula:
\begin{description}
\item[$\posop(\phi, L)$:] Remove tautologies, as well as clauses satisfied by the literals: $\posop(\phi, L) = \{ C \in \phi \mid  C \cap L \not = \emptyset\}$.

@@

\begin{itemize}
\item
The use of negation can be optimized as follows:
\begin{itemize}
\item Constants: $\neg \tautology = \nil$ and $\neg \nil = \tautology$
\item Literals: $\neg x = \obar{x}$ and $\neg \obar{x} = x$
\item Negations: $\neg \neg \psi = \psi$
\end{itemize}

@@

@@

\item[Literals:]
  $\traverse(\lit, \phi) = \lit$ if $\dependencyset(\lit) \subseteq \dvarset$, and $\traverse(\lit, \phi) = \tautology$ if $\dependencyset(\lit) \in \pvarset$.
\item[Product Operations:]
  Let the product operation be of the form $P(\lit_1, \lit_2, \ldots, \lit_k, \psi_1, \psi_2, \ldots, \psi_m)$, where each $\psi_i$ is either a sum or a  product operation.
  \begin{enumerate}
  \item Let $L = \{\lit_1, \lit_2, \ldots, \lit_k\}$.
  \item Define clausal formula $\phi'$ to be $\bcp(\simplify{\phi}{L})$
  \item For $1 \leq i \leq m$, recursively compute $\psi'_i = \traverse(\psi_i, \partition{\phi'}{\dependencyset(\psi_i)})$.
  \item Return the product (with simplifications) for the set of arguments $\{\lit \in L \mid  \dependencyset(\lit) \subseteq \dvarset\} \cup \{ \psi'_i \mid  1 \leq i \leq m \}$.
  \end{enumerate}
\item[Sum Operations:]
  Let the sum operation be of the form $S(\lit_1, \lit_2)$ and having decision variable $x$.
  \begin{enumerate}
  \item For $1 \leq i \leq 2$, recursively compute $\psi'_i = \traverse(\psi_i, \phi)$.
  \item If $x \in \dvarset$, then return the sum (with simplifications) $S(\psi'_1, \psi'_2)$.
  \item Otherwise, let $\psi' = \pkc(\ureduce{\phi}{x})$
    \begin{enumerate}
    \item If $\psi' = \nil$, then return the sum (with simplifications) $S(\psi'_1, \psi'_2)$
    \item Otherwise return nested formula $S(\neg S(\neg \psi'_1, \psi'), \psi'_2)$.
    \end{enumerate}
  \end{enumerate}
\end{description}

Of these steps, the most interesting ones involve the handling of sum
operations when the decision variable $x$ is to be projected.  The
recursive calls yielded subformulas $\psi'_1$ and $\psi'_2$, representing the projected compilation of $\phi$ when $x$ is assigned $1$ and $0$, respectively.
These
formulas potentially have overlapping models and so care must be taken
in forming their sum.  Formula $\ureduce{\phi}{x}$ is the clausal
representation of the universal quantification of $\phi$ with respect
to $x$, and so $\psi' = \pkc(\ureduce{\phi}{x})$ will be a POF whose
data models are the intersection of those for $\psi_1'$ and $\psi_2'$.  In
the event that $\psi' = \nil$, the two formulas are disjoint.  This
will occur, for example, when $x$ is a Tseitin variable encoding some
Boolean formula over the data variables.  In that case, the sum can be
formed directly.  If $\psi' \not = \tautology$, we want to
``subtract'' the overlapping models from one of the arguments before
forming the sum with the other.  We can do so using DeMorgan's Laws.  
That is formula $\neg S(\neg \psi'_1, \psi')$ will have as data models $\dmodelset(\psi'_1) - \dmodelset(\psi')$.  Furthermore, the arguments to this sum operation will have disjoint models
since $\dmodelset(\psi') \subseteq \dmodelset(\psi'_1)$, and therefore
$\dmodelset(\psi') \cap \dmodelset(\neg \psi'_1) = \emptyset$.
As can be seen, we could also reverse the role of the two arguments, computing the result
$S(\psi'_1, \neg S(\neg \psi'_2, \psi'))$.

@@

We can improve the efficiency of the traversal by applying the following optimizations:
\begin{itemize}
\item Only data variables in $\psi$.  If $\dependencyset(\psi) \subseteq \dvarset$, then $\traverse(\psi, \phi) = \psi$.
\item Only data variables in $\phi$.  If $\dependencyset(\phi) \subseteq \dvarset$, then $\traverse(\psi, \phi) = \kc(\phi)$.
\item Only projection variables in $\phi$.  If $\dependencyset(\phi) \subseteq \pvarset$, then $\traverse(\psi, \phi) = \tautology$ if formula $\phi$ is satisfiable and $\nil$ otherwise.
\item Cached result. If $\psi$ is a sum operation and some previous traversal had the same clausal formula $\phi$ as its second argument, then the previously computed result can be reused.\footnote{
Caching for product operations is not, in general valid, since the two recursive calls made when processing a sum operation can be for product operations with the same clausal argument.}
\item Pure literal reduction.
Literal $\lit$ is said to be \emph{pure} in $\phi$ if some clause in $\phi$ contains $\lit$, and no clause in $\phi$ contains  $\obar{\lit}$.
Let $\pure(\phi)$ denote the set of all pure literals in $\phi$, and $\pure_{\pvarset}(\phi) = \{ \lit \in \pure(\phi) \mid  \dependencyset(\lit) \subseteq \pvarset \}$.
If $\pure_{\pvarset}(\phi) \not = \emptyset$, then $\phi$ can be reduced as
$\bcp(\simplify{\phi}{\pure_{\pvarset}(\phi)})$.
\end{itemize}

>>>>

  Let $\varset$ denote a set of Boolean variables, and let $\assign$
  be an \emph{assignment} of truth values to some subset of the
  variables, where $0$ denotes false and $1$ denotes true, i.e.,
  $\assign \colon Y \rightarrow \{0,1\}$ for some $Y  \subseteq \varset$.
We say the assignment is \emph{total with respect to a set of variables} $Y$, when it assigns a value to every $x \in Y$.
  The set of all possible total assignments with respect to a set of variables $Y$ is denoted $\uassign(Y)$.

@@

\begin{definition}[Boolean Formulas]
  The set of Boolean formulas is defined recursively.  Each
  formula $\theta$ has an associated \emph{dependency set}
  $\dependencyset(\theta)  \subseteq \varset$, and a set of \emph{models} $\modelset(\theta)$,
  consisting of assignments that are total with respect to $\varset$ and that satisfy the formula:
  \begin{enumerate}
  \item Constant formulas $\nil$ (logical falsehood) and $\tautology$ (tautology) are Boolean formulas,
    with $\dependencyset(\nil) = \dependencyset(\tautology) = \emptyset$, with $\modelset(\nil) = \emptyset$, and with $\modelset(\tautology) = \uassign(\varset)$.
  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
    and $\modelset(x) = \{\assign \in \uassign(\varset) \mid  \assign(x)=1\}$.
  \item For formula $\theta$, its \emph{negation}, written $\boolnot \theta$ is a Boolean formula,
    with $\dependencyset(\boolnot \theta) = \dependencyset(\theta)$ and $\modelset(\boolnot \theta) = \uassign(\varset) - \modelset(\theta)$.
  \item For formulas $\theta_1, \theta_2, \ldots, \theta_k$, their \emph{product} $\theta = \bigwedge_{1 \leq i \leq k} \theta_i$ is a Boolean formula, with
      $\dependencyset(\theta) = \bigcup_{1 \leq i \leq k} \dependencyset(\theta_i)$ and
      $\modelset(\theta) = \bigcap_{1 \leq i \leq k} \modelset(\theta_i)$.
  \item For formulas $\theta_1, \theta_2, \ldots, \theta_k$, their \emph{sum} $\theta = \bigvee_{1 \leq i \leq k} \theta_i$ is a Boolean formula, with
      $\dependencyset(\theta) = \bigcup_{1 \leq i \leq k} \dependencyset(\theta_i)$ and
      $\modelset(\theta) = \bigcup_{1 \leq i \leq k} \modelset(\theta_i)$.
  \end{enumerate}
\label{def:boolean}
\end{definition}

@@

  Many tools for reasoning about Boolean formulas operate on formulas in \emph{conjunctive normal form}, defined as follows.
  A formula is in  \emph{negation normal form} when negation is applied only to variables.  A
  formula is in \emph{conjunctive normal form} (CNF) when i) it is in
  negation normal form, and ii) sum is applied only to literals.  We use symbol $\phi$ to denote CNF formulas.

  A CNF
  formula can be represented as a set of \emph{clauses}, each of which is a
  set of literals.  Each clause represents the sum of the
  literals, and the formula is the product of its clauses.  We use
  set notation to reference the clauses within a formula and the
  literals within a clause.  A clause consisting of a single literal is referred to as a \emph{unit} clause and the literal as a \emph{unit} literal.
This literal must be assigned value $1$ by any satisfying assignment of the formula.  We use symbol $\phi$ to denote a CNF formula.

@@

We define the following operations for transforming clausal formula $\phi$ into a new clausal formula:
\begin{description}
\item[$\posop(\phi, L)$:] Remove tautologies, as well as clauses satisfied by the literals: $\posop(\phi, L) = \{ C \in \phi \mid  C \cap L = \emptyset\}$.

@@

\begin{itemize}
\item
The use of negation can be optimized as follows:
\begin{itemize}
\item Constants: $\neg \tautology = \nil$ and $\neg \nil = \tautology$
\item Literals: $\neg x = \obar{x}$ and $\neg \obar{x} = x$
\item Double negations: $\neg \neg \psi = \psi$
\end{itemize}

@@

With these simplifications, we can assume that the POFs we generate will satisfy the following:
\begin{itemize}
\item Constant values $\tautology$ and $\nil$ only occur when the POF is itself
  $\tautology$ or $\nil$.
\item Negation is only applied to product and sum operations.
\item Every product and sum operation will have at least two arguments.
\end{itemize}

@@

\item[Literals:]
  $\traverse(\lit, \phi) = \lit$ if $\dependencyset(\lit) \subseteq \dvarset$, and $\traverse(\lit, \phi) = \tautology$ if $\dependencyset(\lit) \in \pvarset$.
\item[Product Operations:]
  Let the product operation be of the form $P(\lit_1, \lit_2, \ldots, \lit_k, \psi_1, \psi_2, \ldots, \psi_m)$, where each $\psi_i$ is either a sum or a  product operation.
  \begin{enumerate}
  \item Let $L = \{\lit_1, \lit_2, \ldots, \lit_k\}$.
  \item Define clausal formula $\phi'$ to be $\bcp(\simplify{\phi}{L})$
  \item For $1 \leq i \leq m$, recursively compute $\psi'_i = \traverse(\psi_i, \partition{\phi'}{\dependencyset(\psi_i)})$.  (Partitioning is only required when $m > 1$.)
  \item Return the product (with simplifications) for the set of arguments $\{\lit \in L \mid  \dependencyset(\lit) \subseteq \dvarset\} \cup \{ \psi'_i \mid  1 \leq i \leq m \}$.
  \end{enumerate}
\item[Sum Operations:]
  Let the sum operation be of the form $S(\lit_1, \lit_2)$ and having decision variable $x$.
  \begin{enumerate}
  \item For $1 \leq i \leq 2$, recursively compute $\psi'_i = \traverse(\psi_i, \phi)$.
  \item If $x \in \dvarset$, then return the sum (with simplifications) $S(\psi'_1, \psi'_2)$.
  \item Otherwise, let $\psi' = \pkc(\ureduce{\phi}{x})$
    \begin{enumerate}
    \item If $\psi' = \nil$, then return the sum (with simplifications) $S(\psi'_1, \psi'_2)$
    \item Otherwise return nested formula $S(\neg S(\neg \psi'_1, \psi'), \psi'_2)$.
    \end{enumerate}
  \end{enumerate}
\end{description}

Of these steps, the most interesting ones involve the handling of sum
operations when the decision variable $x$ is to be projected.  The
recursive calls yield subformulas $\psi'_1$ and $\psi'_2$, representing the projected compilation of $\phi$ when $x$ is assigned $1$ and $0$, respectively.
These
formulas potentially have overlapping models and so care must be taken
in forming their sum.  Formula $\ureduce{\phi}{x}$ is the clausal
representation of the universal quantification of $\phi$ with respect
to $x$, and so $\psi' = \pkc(\ureduce{\phi}{x})$ will be a POF whose
data models are the intersection of those for $\psi_1'$ and $\psi_2'$.  In
the event that $\psi' = \nil$, the two formulas are disjoint.  This
will occur, for example, when $x$ is a Tseitin variable encoding some
Boolean formula over the data variables.  In that case, the sum can be
formed directly.  If $\psi' \not = \tautology$, we want to
``subtract'' the overlapping models from one of the arguments before
forming the sum with the other.  We can do so using DeMorgan's Laws.  
That is formula $\neg S(\neg \psi'_1, \psi')$ will have as data models $\dmodelset(\psi'_1) - \dmodelset(\psi')$.  Furthermore, the arguments to this sum operation will have disjoint models
since $\dmodelset(\psi') \subseteq \dmodelset(\psi'_1)$, and therefore
$\dmodelset(\psi') \cap \dmodelset(\neg \psi'_1) = \emptyset$.
As can be seen, we could also reverse the role of the two arguments, computing the result
$S(\psi'_1, \neg S(\neg \psi'_2, \psi'))$.

@@

We can improve the efficiency of the traversal by applying the following optimizations:
\begin{itemize}
\item Only data variables in $\psi$.  If $\dependencyset(\psi) \subseteq \dvarset$, then $\traverse(\psi, \phi) = \psi$.
\item Only data variables in $\phi$.  If $\dependencyset(\phi) \subseteq \dvarset$, then $\traverse(\psi, \phi) = \kc(\phi)$.
\item Only projection variables in $\phi$.  If $\dependencyset(\phi) \subseteq \pvarset$, then $\traverse(\psi, \phi) = \tautology$ if formula $\phi$ is satisfiable and $\nil$ otherwise.
\item Cached result. If $\psi$ is a sum operation and some previous traversal had the same clausal formula $\phi$ as its second argument, then the previously computed result can be reused.\footnote{
Caching for product operations is not, in general valid, since the two recursive calls made when processing a sum operation can be for product operations with the same clausal argument.}
\item Pure literal reduction.
Literal $\lit$ is said to be \emph{pure} in $\phi$ if some clause in $\phi$ contains $\lit$, and no clause in $\phi$ contains  $\obar{\lit}$.
Let $\pure(\phi)$ denote the set of all pure literals in $\phi$, and $\pure_{\pvarset}(\phi) = \{ \lit \in \pure(\phi) \mid  \dependencyset(\lit) \subseteq \pvarset \}$.
If $\pure_{\pvarset}(\phi) \not = \emptyset$, then $\phi$ can be reduced as
$\phi \leftarrow \bcp(\simplify{\phi}{\pure_{\pvarset}(\phi)})$.
\end{itemize}
